---
title: "mine"
output: html_document
---
# Library
```{r}
library(tidyverse)
library(AlphaSimR)
library(MASS)
library(Matrix)
```



# parameters

So, he selected a bunch of starting parameters.

```{r}
nFounder <- 5
nChr <- 2
segSites <- 10

# WHAT IS THIS?
nSnpPerChr <- 200       # SNPs per chromosome on chip (observed)

```

nSnpPerChr ??? 
I need more explanation


```{r}
## Population / study design
nParents   <- 10      # size of adult generation
nMothers   <- 5       # number of mothers in adult generation
lambdaKids <- 1.6       # mean kids per mother among mothers with >=1 child
```

Here we have the chances of the cmc, the chances of having it while pregnant and the chances of the child having ASD.
```{r}
prevCMC_mother <- 0.10
prevPregCMC    <- 0.50  # calibrated per mother (each mother counts equally)
prevASD_child  <- 0.02
```


Pregnacy effect on Autism
```{r}
betaPreg <- 0.35
```


LIABILITY - THRESHOLD MODEL
```{r}
## ASD liability variance targets (excluding pregnancy term)
var_d <- 0.35   # child direct genetic
var_m <- 0.10   # maternal genetic on child's ASD
var_c <- 0.10   # shared maternal env (per mother)
var_e <- 1 - (var_d + var_m + var_c)
stopifnot(var_e > 0)
```
They are the variance components that the blupf90 should try to guess.



Correlations.
What are we talking about?
How does the math work?
Correlations between things?

```{r}
## Correlations among causal SNP effects across components
rho_CMC_d <- 0.30
rho_CMC_m <- 0.20
rho_d_m   <- 0.10
```




Why do we do this:
```{r}
## Causal architecture
nCausalPerChr <- 50      # causal segregating sites per chromosome
```



GOT IT, STRUCTURE PARTNERSHIPS
```{r}
## Union / partnership structure (divorce/remarriage)
maxPartners_mom <- 3     # max distinct fathers per mother
p_new_partner   <- 0.25  # higher => more mothers with multiple partners
mean_unions_dad <- 1.3   # higher => more paternal half-sibs + higher dad reuse
```


Realistic genotyped fracion
```{r}
## Genotyped fraction for export
propGenotyped <- 0.30
```

# Helper functions
```{r}
## Zero-truncated Poisson with target mean: E[X | X>=1] = mean_target
rztpois_mean <- function(n, mean_target) {
  stopifnot(mean_target > 1)  # zero-truncated Poisson mean is always > 1
  f <- function(mu) mu / (1 - exp(-mu)) - mean_target
  mu0 <- uniroot(f, interval = c(1e-8, 1e4))$root

  x <- rpois(n, mu0)
  while (any(x == 0)) {
    idx <- which(x == 0)
    x[idx] <- rpois(length(idx), mu0)
  }
  x
}

```

Testing it out

```{r}
rztpois_mean_test <- function(n, mean_target) {
  stopifnot(mean_target > 1)  # zero-truncated Poisson mean is always > 1
  f <- function(mu) mu / (1 - exp(-mu)) - mean_target
  
  mu0 <- uniroot(f, interval = c(1e-8, 1e4))$root
  
  print(paste("mu0:", mu0))

  x <- rpois(n, mu0) #SIMULATE POIS RANODM NUMBER WITH MEAN MU0. N NUMBERS EXTRACTED MEAN OF THIS SHOULD BE CLOSE TO MEAN TARGET
  print(paste("x:", x))
  while (any(x == 0)) {
    idx <- which(x == 0)
    print(paste("idx:",idx))
    x[idx] <- rpois(length(idx), mu0)
     
  }
  x
}
```

```{r}
rztpois_mean_test(n = 10, mean_target = 1.6)
```
Okay, I understood and it is used to rerun rpois until they get a number different from 0, which is what we want. The mean of the numbers should be close to the mean target (adjusted mean target using function f).
**Helper function: make correlation matrix always positive**
```{r}
## Ensure correlation matrix is positive definite (for mvrnorm)
make_pd_corr <- function(Sigma, eps = 1e-8) {
  ev <- eigen(Sigma, symmetric = TRUE, only.values = TRUE)$values
  if (min(ev) > eps) return(Sigma)
  as.matrix(nearPD(Sigma, corr = TRUE)$mat)
}

```


**Another Helper Function: calib alpha**
This functions wants to standardize the prevalence of a certain disease in the population despite the lp (genetic predictor score, linear predictor) of the individuals.
Essentially to avoid having unrealistic frequencies in the population.
```{r}
## Calibrate logistic intercept alpha so mean(plogis(alpha + lp)) == target
calib_alpha <- function(target, lp, interval = c(-12, 12)) {
  uniroot(function(a) mean(plogis(a + lp)) - target, interval = interval)$root
  browser()
}
```

```{r}
calib_alpha(0.5, lp=1, interval =c(-1,5))
```

**ANOTHER HELPER FUNCTION**
```{r}
## Split total kids into k blocks (each >=1) for union-based full-sib clusters
alloc_kids_blocks <- function(total_kids, k_blocks) {
  if (k_blocks == 1) return(total_kids)
  x <- rep(1L, k_blocks)
  rem <- total_kids - k_blocks
  if (rem > 0) {
    x <- x + as.integer(rmultinom(1, size = rem, prob = rep(1, k_blocks))[, 1])
  }
  x
}
```

1L --> just means integer values. xL means x is an integer

_I am not sure what k blocks are and what is the purpose of this_ --> for JUAN


# Simulate Pedigrees

First thing we need to simulate the founder haplotypes.

```{r}
## Simulate founder haplotypes
founderPop <- quickHaplo(nInd = 10, nChr= 2, segSites = 10, )
founderPop
```


The founder is MapPop. WHat is this object.
*Ploidy:* 2 means diploid, 4 means tetraploid, etc.
*Loci:* 10 means 10 loci per chromosome, 20 means 20 loci per chromosome, etc.
Why do we have 20 locis? Because we have 2 chromosomes and 10 loci per chromosome, so 2*10 = 20.
So, segSites*nChr is the total number of loci in the genome.

**LETS PLAY WITH MAPPOP**
```{r}
founderPop[1]
```
```{r}
founderPop[11]
```
The number of individuals is the list of the array

```{r}
# Check 
isMapPop(founderPop)
```
```{r}
# can be concatenated
c(founderPop[1], founderPop[2])
```

```{r}
# Slots = Attributes
slotNames(founderPop)
```
```{r}
founderPop <- quickHaplo(nInd = 10, nChr= 2, segSites = 5, )
for (i in slotNames(founderPop)) {
  print(paste("Slot:", i))
  print(slot(founderPop, i))
}
```

**genMAP**

```{r}
slot(founderPop, "genMap")
```

it looks like they all are equally distant to eachother.




